String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-11 2023-1-19/sections/cefa.tex
experiments and resolve the comments of denghang

In this section, we define the model of cost-enriched finite state automata (CEFA), which was introduced in \cite{atva2020} and will be used to solve the satisfiability problem of RECL later on. 
%
Intuitively, CEFA add\denghang{adds} write-only cost registers to finite state automata. Here by ``write-only'', we mean that the cost registers can only be written/updated, but cannot be read, that is, cannot be used to guard the transitions.

@@

The semantics of CEFA is define\denghang{defined} as follows. Let $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$ be a CEFA. 
A \emph{configuration} of $\aut$ is a pair $(q, \vec{v})$ where $q \in Q$ and $\vec{v}$ is a vector denoting the values of registers.  
%An \emph{initial configuration} of $\aut$ is $(q_0, \vec{0})$ with $q_0 \in I$, where the value of each register is zero. 
A \emph{run} of $\aut$ on a string $w = a_1 \cdots a_n$ is a sequence $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ such that $q_0 \in I$ and $q_{i-1} \xrightarrow[\myvec{v_i}]{a_i} q_i$ for each $i \in [n]$. A run $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ is \emph{accepting} if $q_n \in F$ and $\alpha(\myvec{v'}/R)$ is true, where $\myvec{v'} = \sum \limits_{j \in [n]} \myvec{v_j}$. The vector $\myvec{v'}= \sum \limits_{j \in [n]} \myvec{v_j}$ is called the \emph{cost} of an accepting run $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$. Note here \denghang{that} we assume that the initial values of all registers are zero and $\sum \limits_{j \in [n]} \myvec{v_j}$ is the vector of register values after all the transitions in the run are executed. We use $\myvec{v'} \in \aut(w)$ to denote the fact that there is an accepting run of $\aut$ on $w$ whose cost is $\myvec{v'}$.  
We define the semantics of a CEFA $\aut$, denoted by $\Lang(\aut)$, as $\{(w; \myvec{v'}) \mid  \myvec{v'} \in \aut(w)\}$.  In particular, if $I \cap F \neq \emptyset$, then $(\varepsilon; \myvec{0}) \in \Lang(\aut)$. Moreover, we define the \emph{output} of a CEFA $\aut$, denoted by $\cefaout(\aut)$, as $\{\myvec{v'} \mid  \exists w.\ \myvec{v'} \in \aut(w)\}$.

We would like to remark that the accepting conditions $\alpha$ in CEFA are defined in a \emph{global} fashion in the sense that the accepting condition does not distinguish final states. This technical choice is made so that \denghang{the} determinization and minimization of NFA can be utilized to reduce the size of CEFA in the next section.

>>>>

In this section, we define the model of cost-enriched finite state automata (CEFA), which was introduced in \cite{atva2020} and will be used to solve the satisfiability problem of RECL later on. 
%
Intuitively, CEFA add write-only cost registers to finite state automata. Here by ``write-only'', we mean that the cost registers can only be written/updated, but cannot be read, that is, cannot be used to guard the transitions.

@@

The semantics of CEFA is defined as follows. Let $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$ be a CEFA. 
A \emph{configuration} of $\aut$ is a pair $(q, \vec{v})$ where $q \in Q$ and $\vec{v}$ is a vector denoting the values of registers.  
%An \emph{initial configuration} of $\aut$ is $(q_0, \vec{0})$ with $q_0 \in I$, where the value of each register is zero. 
A \emph{run} of $\aut$ on a string $w = a_1 \cdots a_n$ is a sequence $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ such that $q_0 \in I$ and $q_{i-1} \xrightarrow[\myvec{v_i}]{a_i} q_i$ for each $i \in [n]$. A run $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ is \emph{accepting} if $q_n \in F$ and $\alpha(\myvec{v'}/R)$ is true, where $\myvec{v'} = \sum \limits_{j \in [n]} \myvec{v_j}$. The vector $\myvec{v'}= \sum \limits_{j \in [n]} \myvec{v_j}$ is called the \emph{cost} of an accepting run $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$. Note that we assume that the initial values of all registers are zero and $\sum \limits_{j \in [n]} \myvec{v_j}$ is the vector of register values after all the transitions in the run are executed. We use $\myvec{v'} \in \aut(w)$ to denote the fact that there is an accepting run of $\aut$ on $w$ whose cost is $\myvec{v'}$.  
We define the semantics of a CEFA $\aut$, denoted by $\Lang(\aut)$, as $\{(w; \myvec{v'}) \mid  \myvec{v'} \in \aut(w)\}$.  In particular, if $I \cap F \neq \emptyset$, then $(\varepsilon; \myvec{0}) \in \Lang(\aut)$. Moreover, we define the \emph{output} of a CEFA $\aut$, denoted by $\cefaout(\aut)$, as $\{\myvec{v'} \mid  \exists w.\ \myvec{v'} \in \aut(w)\}$.

We would like to remark that the accepting conditions $\alpha$ in CEFA are defined in a \emph{global} fashion in the sense that the accepting condition does not distinguish final states. This technical choice is made so that the determinization and minimization of NFA can be utilized to reduce the size of CEFA in the next section.

### 2023-05-11 2023-1-19/sections/implementation.tex
experiments and resolve the comments of denghang

We implement the algorithms in Section~\ref{sec:algorithm} on top of OSTRICH~\cite{ostrich,atva2020}, resulting into\denghang{in} a version of OSTRICH called OSTRICH$^{\rm RECL}$.
Then we do experiments on two benchmark suites, to evaluate the performance of OSTRICH$^{\rm RECL}$. In the sequel, we first describe the two benchmark suites, then present the settings and results of the experiments.

@@

Q1 is answered by running OSTRICH$^{\rm RECL}$ and the aforementioned solvers on AutomatArk and RegCoL. The experimental results can be found in Table~\ref{tab:results_all}. From Table~\ref{tab:results_all}, we can see that OSTRICH$^{\rm RECL}$ solvers almost all 48,834 instances, except 144 of them, that is, it solves 48,689 instances correctly, which is 3,982/2,263/24,568/12,823/1,082/7083 more than the number of instances solved by CVC5/Z3seq/Z3-Trau/Z3str3/Z3str3RE/OSTRICH respectively. 
%    
Note that Z3seq, Z3-Trau, Z3str3, and Z3str3RE have soundness errors, where the result of CVC5 is taken as the ground truth. If CVC5 reports ``unknown'' or ``timeout'' in some instance, then no soundness errors will be reported in this instance. 
%A soundness error is reported if the result of a solver for an instance is different from CVC5. 
% their results on some instances are inconsistent with the results of the other three solvers, i.e. CVC5, OSTRICH, and OSTRICH$^{\rm RECL}$.
%
Moreover, OSTRICH$^{\rm RECL}$ is the second fastest solver. More specifically, OSTRICH$^{\rm RECL}$ spends 2.93 seconds per instance on average, while the fastest solver Z3str3RE spends 1.71 seconds per instance, and the other solvers spend at least 3 seconds per instance.  
Z3-Trau reports ``unknown'' on 21,655 instances, which is mainly attributed to the fact that it does not support \verb|re.diff|, the language difference operator of regular expressions.

@@

To answer Q2, we compare OSTRICH$^{\rm RECL}$ with its three variants, namely, OSTRICH$^{\rm RECL}_{\rm -SR}$, OSTRICH$^{\rm RECL}_{\rm -UA}$, OSTRICH$^{\rm RECL}_{\rm -SRUA}$, which remove size-reduction, under-approximation, and both from OSTRICH$^{\rm RECL}$ respectively. The experiment results can be found in Table~\ref{tab:results_heuristics}.
From Table~\ref{tab:results_heuristics}, we can see that adding size-reduction techniques help solve \zhilin{xxx} more instances and reduce the average time per instance for \zhilin{xxx}/\%, while adding under-approximation techniques help solve \zhilin{xxx} more instances and reduce the average time per instance for \zhilin{xxx}/\%, while adding both help solve \zhilin{xxx} more instances and reduce the average time per instance for \zhilin{xxx}/\%. 
%Moreover, adding these techniques also decrease the average time spent in solving each instance considerably.

@@

\zhilin{stopped here}

>>>>

We implement the algorithms in Section~\ref{sec:algorithm} on top of OSTRICH~\cite{ostrich,atva2020}, resulting in a version of OSTRICH called OSTRICH$^{\rm RECL}$.
Then we do experiments on two benchmark suites, to evaluate the performance of OSTRICH$^{\rm RECL}$. In the sequel, we first describe the two benchmark suites, then present the settings and results of the experiments.

@@

Q1 is answered by running OSTRICH$^{\rm RECL}$ and the aforementioned solvers on AutomatArk and RegCoL. The experimental results can be found in Table~\ref{tab:results_all}. From Table~\ref{tab:results_all}, we can see that OSTRICH$^{\rm RECL}$ solvers almost all 48,834 instances, except 145 of them, that is, it solves 48,698 instances correctly, which is 3,994/2,272/24,577/12,832/1,091/7,092 more than the number of instances solved by CVC5/Z3seq/Z3-Trau/Z3str3/Z3str3RE/OSTRICH respectively. 
%    
Note that Z3seq, Z3-Trau, Z3str3, and Z3str3RE have soundness errors, where the result of CVC5 is taken as the ground truth. If CVC5 reports ``unknown'' or ``timeout'' in some instance, then no soundness errors will be reported in this instance. 
%A soundness error is reported if the result of a solver for an instance is different from CVC5. 
% their results on some instances are inconsistent with the results of the other three solvers, i.e. CVC5, OSTRICH, and OSTRICH$^{\rm RECL}$.
%
Moreover, OSTRICH$^{\rm RECL}$ is the second fastest solver. More specifically, OSTRICH$^{\rm RECL}$ spends 2.93 seconds per instance on average, while the fastest solver Z3str3RE spends 1.71 seconds per instance, and the other solvers spend at least 3.49 seconds per instance.  
Note that Z3-Trau reports ``unknown'' on 21,655 instances, which seems to be mainly attributed to the fact that it does not support \verb|re.diff|, the language difference operator of regular expressions.

@@

To answer Q2, we compare OSTRICH$^{\rm RECL}$ with its three variants, namely, OSTRICH$^{\rm RECL}$ where the size-reduction, under-approximation, or both are removed. 
For convenience, let us use OSTRICH$^{\rm RECL^-}$ to denote the variant of OSTRICH$^{\rm RECL}$ that both size-reduction and under-approximation techniques are removed, 
OSTRICH$^{\rm RECL^-}_{+\rm SR}$ (resp. OSTRICH$^{\rm RECL^-}_{+\rm UA}$) to denote the variant that is obtained by adding the size-reduction (resp. under-approximation) techniques to OSTRICH$^{\rm RECL^-}$.
%OSTRICH$^{\rm RECL}_{\rm -SR}$, OSTRICH$^{\rm RECL}_{\rm -UA}$, OSTRICH$^{\rm RECL}_{\rm -SRUA}$, which remove size-reduction, under-approximation, and both from OSTRICH$^{\rm RECL}$ respectively. 
The experiment results can be found in Table~\ref{tab:results_heuristics}.
From Table~\ref{tab:results_heuristics}, we can see that adding size-reduction techniques help to solve 1,563 more instances and reduce the average time per instance for 46.4\% (see the OSTRICH$^{\rm RECL^-}_{\rm +SR}$ column), while adding under-approximation techniques help to solve 1,476 more instances and reduce the average time per instance for 42.7\% (see the OSTRICH$^{\rm RECL^-}_{\rm +UA}$ column), moreover, adding both help to solve 1,579 more instances and reduce the average time per instance for 46.5\% (see the OSTRICH$^{\rm RECL}$ column). Note that adding under-approximation to OSTRICH$^{\rm RECL^-}_{\rm +SR}$ help to solve 19 more sat instances, while reduces the number of solved unsat instances by 3, since under approximation techniques do not help to solve unsat instances.
%Moreover, adding these techniques also decrease the average time spent in solving each instance considerably.

@@

%\zhilin{stopped here}

### 2023-05-11 2023-1-19/sections/introduction.tex
experiments and resolve the comments of denghang

In modern programming languages---such as JavaScript, Python, Java, and PHP---the string data type plays a crucial role. 
It is well-known that string manipulations are error-prone and could even give rise to security vulnerabilities (e.g. \ cross-site scripting, aka XSS). 
One powerful method for identifying such bugs in programs is \emph{symbolic execution} (possibly in combination with dynamic analysis), which analyses symbolic paths in a program by viewing them as constraints, whose feasibility is checked by constraint solvers. 
%
Symbolic execution of string manipulating programs has motivated the highly active research area of \emph{string constraint solving}, resulting in the development of numerous string solvers in the last decade or so\denghang{or so?}, including the state-of-the-art string constraint solvers (e.g.
Z3~\cite{Z3}, CVC4/5~\cite{cvc4}, Z3str/2/3/4~\cite{Z3-str,Z3-str2,Z3-str3,BerzishMurphy2021}, Z3str3RE~\cite{BD+23}, 
Z3-Trau~\cite{Z3-trau}, OSTRICH~\cite{CHL+19}, Slent~\cite{WC+18}, among many others.

@@

The representation of regex-counting and string-length as registers in CEFA, combined with the aforementioned size-reduction and under-approximation techniques, entails an efficient algorithm for solving RECL constraints. We implement the algorithm on top of OSTRICH, resulting in a new version of OSTRICH, called $\ostrichrecl$ for convenience. 
%
Furthermore, we utilize a collection of benchmark suites, which comprise 49,379 RECL constraints in total, to evaluate the performance of $\ostrichrecl$. The collection of benchmark suites includes a widely-used benchmark suite for string constraints, as well as benchmark suites comprising constraints generated from regular expressions in various open source regex libraries. 
%
%Note that in the benchmark suites, we only consider the regexes that contain explicit counting, that is, regexes that contain sub-expressions of the form $e^{\{n, m\}}$, and ignore those that contain only implicit counting operators, that is, $e^*$ or $e^+$, since the state-of-the-art string solvers suffer mainly from the explicit counting operators. 
%
The experimental results show that $\ostrichrecl$ can solve 49,128 out of 49,379 instances, while the state-of-the-art best solver Z3str3RE solves only 47,407 instances, with the timeout set as 60 seconds. Moreover, $\ostrichrecl$ and Z3str3RE spend almost the same amount of time, on average, 2.5 seconds, to solve each instance. 
In particular, the string constraint in~(\ref{eqn-running}) can be solved by $\ostrichrecl$ within one second. 
In comparison, CVC5, Z3str3, and Z3seq solve only 44,001, 36,320, 46,813 instances respectively, with an average time 7.8, 7.7, 3.8 seconds per instance.

At last, we would like to clarify the relationship of this work with~\cite{atva2020}, where CEFA was introduced. This work differs from \cite{atva2020} in the following aspects: In \cite{atva2020}, the regex-counting was unfolded explicitly. Moreover, in~\cite{atva2020}, the nonemptiness of CEFA was solved by utilizing the model checker nuXmv~\cite{nuxmv}. In contrast, this work solves it by integrating the techniques of Parikh image computation, automata-size reduction, under approximation, and SMT solvers for LIA. The experimental results show that our approach achieves better performance than that in~\cite{atva2020}. \zhilin{this point should be supposed by experimental results.}

>>>>

In modern programming languages---such as JavaScript, Python, Java, and PHP---the string data type plays a crucial role. 
It is well-known that string manipulations are error-prone and could even give rise to security vulnerabilities (e.g. \ cross-site scripting, aka XSS). 
One powerful method for identifying such bugs in programs is \emph{symbolic execution} (possibly in combination with dynamic analysis), which analyses symbolic paths in a program by viewing them as constraints, whose feasibility is checked by constraint solvers. 
%
Symbolic execution of string manipulating programs has motivated the highly active research area of \emph{string constraint solving}, resulting in the development of numerous string solvers in the last decade, including the state-of-the-art string constraint solvers (e.g.
Z3~\cite{Z3}, CVC4/5~\cite{cvc4}, Z3str/2/3/4~\cite{Z3-str,Z3-str2,Z3-str3,BerzishMurphy2021}, Z3str3RE~\cite{BD+23}, 
Z3-Trau~\cite{Z3-trau}, OSTRICH~\cite{CHL+19}, Slent~\cite{WC+18}, among many others.

@@

The representation of regex-counting and string-length as registers in CEFA, combined with the aforementioned size-reduction and under-approximation techniques, entails an efficient algorithm for solving RECL constraints. We implement the algorithm on top of OSTRICH, resulting in a new version of OSTRICH, called $\ostrichrecl$ for convenience. 
%
Furthermore, we utilize a collection of benchmark suites, which comprise 48,834 RECL constraints in total, to evaluate the performance of $\ostrichrecl$. The collection of benchmark suites includes a widely-used benchmark suite for string constraints, as well as benchmark suites comprising constraints generated from regular expressions in various open source regex libraries. 
%
%Note that in the benchmark suites, we only consider the regexes that contain explicit counting, that is, regexes that contain sub-expressions of the form $e^{\{n, m\}}$, and ignore those that contain only implicit counting operators, that is, $e^*$ or $e^+$, since the state-of-the-art string solvers suffer mainly from the explicit counting operators. 
%
The experimental results show that $\ostrichrecl$ can solve 48,698 out of 48,834 instances, while the state-of-the-art best solver Z3str3RE on RECL constraints solves only 47,607 instances, with the timeout period set as 60 seconds. In particular, the string constraint in~(\ref{eqn-running}) can be solved by $\ostrichrecl$ within one second. 
Moreover, while $\ostrichrecl$ solves 1,091 more instances than Z3str3RE, it is only slightly slower than it (2.93 versus 1.71 seconds per instance). 
In comparison, CVC5, Z3seq, Z3-Trau, and Z3str3 solve only 44,704, 46,426, 24,121, and 35,866 instances respectively, with an average time 5.67, 3.49, 5.41, and 8.54 seconds per instance. (See Table~\ref{tab:results_all})

At last, we would like to clarify the relationship of this work with~\cite{atva2020}, where CEFA was introduced. This work differs from \cite{atva2020} in the following aspects: In \cite{atva2020}, the regex-counting was unfolded explicitly. Moreover, in~\cite{atva2020}, the nonemptiness of CEFA w.r.t. LIA formulas was solved by utilizing the model checker nuXmv~\cite{nuxmv}. In contrast, this work solves it by integrating the techniques of Parikh image computation, automata-size reduction, and under approximation. The experimental results show that our approach solves 7,092 more instances while spending less time than the solver in~\cite{atva2020}. (See Table~\ref{tab:results_all}, where OSTRICH denotes the solver in \cite{atva2020}.)
